---
title: "Creating Accounts and Topics with Hadeda"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating Accounts and Topics with Hadeda}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = interactive()
)
```

This vignette walks through creating a Hedera account and a Hedera Consensus Service topic using the `hadeda` gRPC helpers. It assumes you have an operator account (ID + private key) funded on the Hedera testnet and that the required native gRPC libraries are installed. See `vignettes/hadeda-grpc-setup.Rmd` or the project README for environment setup details.

> **Heads up**  
> These examples submit live transactions and incur fees. Run them against testnet unless you explicitly intend to interact with mainnet.

## 1. Load credentials

`hadeda` looks for operator details via environment variables. You can set them before launching R:

```bash
export HADEDA_NETWORK=testnet
export HADEDA_OPERATOR_ID=0.0.1234
export HADEDA_OPERATOR_KEY=302e0201...
```

Alternatively, configure them in-session:

```{r auth, eval = FALSE}
library(hadeda)

config <- hadeda_auth_config(
  network      = "testnet",
  operator_id  = Sys.getenv("HADEDA_OPERATOR_ID"),
  operator_key = Sys.getenv("HADEDA_OPERATOR_KEY")
)

# Merge your gRPC handler definitions (see hadeda-grpc-setup vignette)
# config$grpc <- modifyList(config$grpc, grpc_handler)
```

The helper stores credentials in-memory for subsequent gRPC calls.

> **Security tip**  
> Store long-lived keys in an OS-backed vault via [`keyring`](https://cran.r-project.org/package=keyring)
> (`keyring::key_set_with_value()` / `keyring::key_get()`) rather than plain-text `.Renviron` files. For temporary prompts,
> combine `keyring` with [`askpass`](https://cran.r-project.org/package=askpass). Use `withr::with_envvar()` to inject values
> for automated scripts without leaking them to the global session.

## 2. Generate a key for the new account

Accounts must be created with a public key. Generate an ed25519 keypair:

```{r generate-key, eval = FALSE}
new_key <- hadeda_crypto_generate_private_key()
pub_key <- hadeda_crypto_public_key(new_key)
```

Persist the private key securely if you plan to reuse this account later—`keyring` can store it under a descriptive service
name (for example `hadeda/testnet-account-private-key`) and retrieve it only when required.

## 3. Create the account

Call `hadeda_crypto_create_account()` to submit the transaction. Initial balance is denominated in tinybars.

```{r create-account, eval = FALSE}
new_account <- crypto_create_account(
  config,
  public_key      = pub_key,
  initial_balance = 10  # 10 tinybars
)

new_account$id
```

The returned tibble includes the new account ID and receipt metadata.

## 4. Create a topic

With the operator configured, creating a consensus topic is straightforward:

```{r create-topic, eval = FALSE}
new_topic <- topics_create(
  config,
  memo = "Demo topic created via hadeda vignette",
  .transport = "rest"  # topic creation currently routes through REST
)

new_topic$id
```

The REST transport targets Hashio's Mirror Node by default; set `config$rest$base_url` and headers (for example an `X-API-Key`) if you use a different provider. Topics can be further customized by specifying admin or submit keys. Refer to `?topics_create` for the full argument set.

## 5. Next steps

With the account and topic in place, you can:

- Transfer HBAR or tokens using `hadeda_crypto_transfer()` and token helpers.
- Submit messages to the topic with `hadeda_consensus_submit_message()`.
- Manage topic permissions by rotating admin and submit keys.

Explore the other vignettes (`hadeda-tokens.Rmd`, `hadeda-transactions.Rmd`) for extended workflows. Remember to clean up any testnet state you no longer need to avoid unnecessary fees.
